Analyses

### 2024-07-10 manuscript.RmD
- updated manuscript - removed typos - extended theory

abstract          : "The privacy calculus model states that people weigh risks and benefits before communicating online. This paper analyzes three questions. How robust is the privacy calculus, can it be replicated with actual behavior in realistic online settings? Do we need to extend the model by integrating more socio-psychological measures such as trust, self-efficacy, and privacy deliberation? How easily can the privacy calculus be manipulated by external factors such as like and dislike buttons? In a preregistered one-week field experiment (N = 590), participants discussed a current political topic on an online website, which included either (a) like buttons, (b) both like and dislike buttons, or (c) no buttons. The privacy calculus model was confirmed: Benefits and concerns affected communication. Deliberating about privacy by comparing benefits and risks decreased communication. Experiencing self-efficacy and trust increased communication. Like and dislike buttons had a strong effect on behavior: They reduced communication by 45 percent."

@@

bibliography      :
  - [bibliography/r-references.bib]
  - [bibliography/references.bib]
floatsintext      : yes
figsintext        : yes
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : yes
wordcount         : "8,187"
draft             : no
# output            : papaja::apa6_pdf
# documentclass     : "apa6"
# header-includes:
#   - \setlength{\parskip}{0em}
#   - \raggedbottom
#   - \note{\clearpage}
# output            : papaja::apa6_pdf
# documentclass     : "apa7"
# classoption       : "man"
# header-includes:
#   - |
#     \makeatletter
#     \renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
#       {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
#       {-1em}%
#       {\normalfont\normalsize\bfseries\typesectitle}}
# 
#     \renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
#       {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
#       {-\z@\relax}%
#       {\normalfont\normalsize\bfseries\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
#     \makeatother
#     \usepackage{endnotes}
#     \setlength{\parskip}{0em}
#     \raggedbottom
#     \note{\clearpage}
# 
# csl               : "`r system.file('rmd', 'apa7.csl', package = 'papaja')`"
# output            : papaja::apa6_docx
output:
  html_document:
    toc: TRUE
    toc_float:
      collapsed: FALSE
      toc_depth: 2
---

@@

invisible(
  lapply(
    pcks, 
    library, 
    character.only = TRUE
    )
  )

@@

According to the privacy paradox, the way users share information online is erratic [@barnesPrivacyParadoxSocial2006]:
People communicate much personal information online, although being strongly concerned about their privacy [@taddickenUsesPrivacyOnline2011].
However, despite its popularity in both academic research [@masurUnderstandingEffectsConceptual2023] and the public press [@newyorkpublicradioPrivacyParadox2018], empirical support for the privacy paradox is tenuous [@baruhOnlinePrivacyConcerns2017; @dienlinDoesPrivacyParadox2021]. 
Research is increasingly building on the privacy calculus model, which states that communication online can be explained by perceived risks and expected benefits [@bolUnderstandingEffectsPersonalization2018; @krasnovaOnlineSocialNetworks2010].

Although the privacy calculus has gained momentum in academic research, several important questions remain unanswered. 
First, the privacy calculus' empirical foundation needs to be improved [@knijnenburgDeathPrivacyCalculus2017]. 
To date, much research on the privacy calculus builds on (a) self-reports of behavior [@krasnovaOnlineSocialNetworks2010], (b) vignette approaches [@bolUnderstandingEffectsPersonalization2018], or (c) one-shot experiments in the lab [@treptePrivacyCalculusContextualized2020]. 
Self-reports are often unreliable measures of online behavior [@parrySystematicReviewMetaanalysis2021], vignette approaches have limited external validity, and one-shot experiments cannot analyze compound effects of sustained use.
The first aim is therefore to replicate the privacy calculus model in a more authentic and long-term setting using actual behavioral data.
Doing so will show its robustness, and it will help understand if behavior is better explained by the privacy paradox approach or by the privacy calculus model.

Second, the privacy calculus model is criticized for neglecting the actual process of weighing pros and cons, and for over-emphasizing rationality  [@knijnenburgDeathPrivacyCalculus2017].
Simply showing that both concerns and gratifications correlate with communication online does not prove that an explicit weighing process took place.
Instead, we need to find out if people deliberately compare costs and benefits before communicating online. 
Therefore, I will also analyze the process of comparing pros and cons explicitly, by elaborating on what I will call the the privacy deliberation process [@omarzuDisclosureDecisionModel2000]. 
Next, focusing exclusively on costs and benefits does not fully capture how and why users engage online.
I, hence, extend the privacy calculus by analyzing how online communication is affected by trust and self-efficacy, two variables less focused on rationality and more oriented toward socio-psychological aspects [@metzgerPrivacyTrustDisclosure2004].

Online behavior is determined by the person itself and their external circumstances [@bazarovaIntegrationIndividualisticNetworked2020; @treptePrivacyCalculusContextualized2020; @spottswoodShouldShareThat2017].
The privacy calculus upholds that much of behavior is rational and self-determined, but how easily can this process be manipulated by external circumstances?
Online contexts are optimized to elicit as much interaction as possible [@ellisonSocialNetworkSite2015; @masurBehavioralContagionSocial2021], for example via low threshold communication features such as likes, shares, replies, or reactions  [@carrPredictingThresholdPerceived2018]. 
Implicit and explicit cues on how a website is used can increase self-disclosure [@spottswoodShouldShareThat2017; @treptePrivacyCalculusContextualized2020].
How large are the effects of external popularity cues such as like and dislike buttons compared to the internal weighing of pros and cons?
And how easily can the privacy calculus be manipulated by popularity cues?

@@

The privacy calculus analyzes why people communicate online.
When are we willing to engage in a conversation?
It builds on the calculus of behavior [@lauferPrivacyConceptSocial1977] and states that people are weighing risks and benefits before engaging in communication.
Communication is closely related to privacy, which is defined as a "voluntary and temporary withdrawal of a person from the general society" [@westinPrivacyFreedom1967, p. 7].
People regulate their privacy by deciding what and what not to communicate [@dienlinPrivacyProcessModel2014].
Sharing information carries risks, as recipients may reuse it in different contexts, potentially harming the original sender [@masurBehavioralContagionSocial2021].
As a result, people who aim to avoid risks should be included to share less information, especially in online contexts where audiences are much larger [@vitakImpactContextCollapse2012].
Indeed, empirical research confirms that people with higher privacy needs and privacy concerns are less likely to communicate online [@krasnovaOnlineSocialNetworks2010; @masurTransformativeNotHow2021; @masurUnderstandingEffectsConceptual2023].
This finding is now confirmed by several meta-analyses and reviews [@baruhOnlinePrivacyConcerns2017; @dienlinDoesPrivacyParadox2021; @gerberExplainingPrivacyParadox2018].
Similarly, people who are more concerned about their privacy also engage in more privacy protection behaviors [@baruhOnlinePrivacyConcerns2017; @stubenvollLivingEasyEyes2022].

Outweighing concerns, the most relevant drivers of online communication are expected gratifications [@dienlinExtendedPrivacyCalculus2016; @bolUnderstandingEffectsPersonalization2018; @kezerGettingPrivacyCalculus2022].
People are happy to trade in parts of their privacy when they can receive something more valuable in return [@lauferPrivacyConceptSocial1977].
In online communication, the most important benefits include social support, social capital, entertainment, information-seeking, and self-presentation [@dhirUnderstandingRelationshipIntensity2017; @ellisonBenefitsFacebookFriends2007; @krasnovaOnlineSocialNetworks2010; @whitingWhyPeopleUse2013].

@@

Although privacy calculus implies that people explicitly compare benefits and risks before communicating online, prior research has neglected this aspect [@knijnenburgDeathPrivacyCalculus2017].
Only observing that privacy concerns or expected gratifications and communication online are related does not prove that an explicit weighing process took place [@knijnenburgDeathPrivacyCalculus2017]. 
Instead, we need to analyze if, and if so by how much, people actively deliberate about their privacy, and whether doing so influences their willingness to communicate.
Self-disclosure theory [@omarzuDisclosureDecisionModel2000;  @altmanPrivacyConceptualAnalysis1976] suggests that if the benefits of communication are attractive, deciding whether or not to communicate is a "conscious and deliberative process"  [@omarzuDisclosureDecisionModel2000, p. 183].
I hence introduce and investigate a novel concept termed _privacy deliberation_.
Privacy deliberation captures the extent to which individuals explicitly compare risks and benefits before communicating with others.

@@

<!-- The aforementioned weighing process is not flawless or perfect. -->
It is useful to understand the privacy calculus from the perspective of _bounded rationality_ [@simonBoundedRationality1990].
Bounded rationality states that "(1) humans are cognitively constrained; (2) these constraints impact decision making; and (3) difficult problems reveal the constraints and highlight their significance." [@bendorBoundedRationality2015, p. 1303].
Although human behavior is considered partly irrational, bounded rationality does not state that it is completely irrational [@gigerenzerBoundedRationalityAdaptive2002].
Rationality can hence be understood to exist on a continuum. 
In the context of online privacy, rationality is impeded by information asymmetries, presence bias, intangibility, illusory control, or herding [@acquistiSecretsLikesDrive2020].
It follows that to provide a more complete picture, additional factors next to gratifications and concerns should also explain communication.
Factors less focused on rationality, but more on socio-psychological aspects.

The two factors perhaps most important when it comes to understanding online communication are self-efficacy and trust [@hossainICTEnabledVirtual2004; @metzgerPrivacyTrustDisclosure2004].
Privacy violations create psychological distress [@ledbetterParentchildPrivacyBoundary2019], experiencing online contexts as a safe space that users can sufficiently control is important for engaging in online communication. 
If users are more familiar, experienced, and knowledgeable in a given online context, they are more likely to navigate that online contexts successfully and to communicate actively [@baruhOnlinePrivacyConcerns2017; @kramerMasteringChallengeBalancing2020; @parkDigitalLiteracyPrivacy2013a].
People with more privacy self-efficacy engage more successfully in self-withdrawal [@dienlinExtendedPrivacyCalculus2016] and protective online behavior [@boermanExploringMotivationsOnline2021a; @vanooijenPrivacyCynicismIts2022].
Hence, if users possess more self-efficacy to participate, they should also communicate more, and to more fully capture online behavior self-efficacy should hence be included.

@@

In all situations where people lack experience, control, or competence, a central variable to understand behavior is trust [@gefenTrustTAMOnline2003].
Especially in online contexts, trust plays a key role [@metzgerPrivacyTrustDisclosure2004; @metzgerEffectsSiteVendor2006a].
Users often cannot control the environment or the way their information is handled [@acquistiSecretsLikesDrive2020; @braunlichLinkingLooseEnds2020].
Trust either captures "_specific_ beliefs dealing primarily with the integrity, benevolence, and ability of another party" [@gefenTrustTAMOnline2003, p. 55, emphasis added] or a "_general_ belief that another party can be trusted" [@gefenTrustTAMOnline2003, p.55, emphasis added]. 
In online contexts, there are different targets of trust, including (a) the information system, (b) the provider, (c) the Internet, and (d) the community of other users [@sollnerWhyDifferentTrust2016]. 
People who put more trust in the providers of networks, for example, disclose more personal information [@liEmpiricalStudiesOnline2011]. 
To comprehensively capture and understand online communication, trust should hence be included.

@@

How are the privacy calculus and communication affected by the context, the digital infrastructure? 
How easily can the calculus be manipulated externally?
One of the central tools to afford and govern online behavior are popularity cues such as like and dislike buttons [@stsiampkouskayaNotExperimentalStudy2023].
Popularity cues have been shown to affect behavior [@kramerMasteringChallengeBalancing2020; @masurBehavioralContagionSocial2021; @treptePrivacyCalculusContextualized2020]:
Online comments that already have several dislikes are more likely to receive further dislikes [@muchnikSocialInfluenceBias2013]. 
When users disagree with a post, they are more likely to click on a button labeled _respect_ compared to a button labeled _like_ [@stroudRecommendRespectAltering2017].

How and why might popularity cues affect the privacy calculus?
In analyzing this question, it makes sense to analyze the cues' underlying affordances [@ellisonSocialNetworkSite2015; @foxDistinguishingTechnologiesSocial2017].
Affordances are mental representations of how objects might be used.
They emphasize that it is not the _objective features_ that determine behavior, but rather our _subjective perceptions_ [@gibsonEcologicalApproachVisual2015].
Popularity cues such as like and dislike buttons, which are of primary interest in this study, are understood as "paralinguistic digital affordances" [@carrPredictingThresholdPerceived2018, p. 142].

```{r model, fig.cap="Overview of analyzed model.", fig.align='center', out.width=".8\\textwidth"}
# knitr::include_graphics("figures/design/model_simplified.pdf")
knitr::include_graphics("figures/design/model_simplified.png")
```

Popularity cues likely impact the privacy calculus via two underlying theoretical mechanisms [@kramerMasteringChallengeBalancing2020]:
First, the _mere presence_ of popularity cues might affect whether people are willing to disclose. 
Being able to attract likes might afford active communication. 
The mere option to receive dislikes, conversely, might inhibit communication.
Second, _actually receiving_ likes or dislikes might affect behavior, by means of positive reinforcement (likes) or punishment (dislikes) [@skinnerScienceHumanBehavior2014].
To illustrate, likes are affirmative and embody the positivity bias of social media [@schreursExposurePositivityBias2022]. 
Receiving a like online is similar to receiving a compliment offline [@carrPredictingThresholdPerceived2018; @sumnerFunctionalApproachFacebook2017].
Like-buttons afford and emphasize a _gain frame_ [@rosoffHeuristicsBiasesCyber2013].
These gains can be garnered only through active participation. 
In situations where people can gain immediate positive outcomes, concerns and risks that are more vague and in the future often become less relevant [@ainslieHyperbolicDiscounting1992].
Because like buttons emphasize positive outcomes, it is likely that concerns decrease. 
In situations where there is more to win, people might also more actively deliberate about whether or not to communicate. 
Dislikes, instead, represent a punishment, introducing a _loss frame_. 
Websites featuring both like _and_ dislike buttons should therefore be more ambivalent compared to websites without popularity cues.
Privacy concerns should not be reduced anymore:
People who are more concerned about their privacy are also more shy and risk averse [@dienlinPsychologyPrivacyAnalyzing2017].
Implementing the dislike button might therefore increase privacy concerns, canceling out the positive effects of the like button.

@@

For a simplified overview of the analyzed model, see Figure \@ref(fig:model).

@@

The data were collected in Germany.
Participants were recruited using the professional panel agency Norstat.
As incentive, participants were awarded digital points, to receive special offers from online retailers. 
Participants had to be above the age of 18 and had to live in Germany. 
In a first step, the company sent its panel members an invitation to participate in the study. 
In this invitation, panel members were asked to participate in a study analyzing the current threat posed by terrorist attacks in Germany.
Members who agreed to participate were sent the first questionnaire (_T1_).
The questionnaire asked participants about their sociodemographic background, (b) provided more details about the study, and (c) included a registration link for the website, which was introduced as "participation platform".
Afterward, participants were randomly assigned to one of the three websites. 
After registration was completed, participants were invited (but not obliged) to visit the website and to discuss the topic of the terrorism threat in Germany. 
Participants could use the website and write comments over the course of one week.
Subsequently, participants received a follow-up questionnaire in which the self-reported measures reported below were collected (_T2_).

@@

```{r CFA, results = "asis"}
apa_table(
  facval_tab %>% 
    select(-alpha) %>% 
    rename(`p-value` = pvalue) %>% 
    mutate(df = as.integer(.$df)), 
  font_size = "footnotesize",
  caption = "Psychometric Properties, Factorial Validity, and Reliability of Measures",
  note = "omega = Raykov's composite reliability coefficient omega; avevar = average variance extracted."
  , align = c("l", rep("r", 11))
  )
```

@@

Two types of gratifications were collected.
_General gratifications_ were measured with five items based on @sunLocationInformationDisclosure2015.
One example item was "Using the participation platform has paid off for me". 
Based on @schererGratifikationMinuteZeitnahe2002, _specific gratifications_ were measured on five dimensions with three items each. 
Example items were: "Using the participation platform made it possible for me to" ... "learn things I would not have noticed otherwise" (information), "react to a subject that is important to me" (relevance), "engage politically" (political participation), "try to improve society" (idealism), and "soothe my guilty consciences" (extrinsic benefits).
The means for both dimensions were _m_ = `r facval_tab["Gratifications general", "m"]` (general gratifications) and _m_ = `r facval_tab["Gratifications specific", "m"]` (specific gratifications), which were above the scales' midpoint.
This shows that on average people considered the website to be beneficial.

@@

I first plotted the bivariate relations of all variables (see Figure \@ref(fig:corrplot)). 
All variables referring to the privacy calculus demonstrated the expected bivariate relationships with communication. 
For example, people who were more concerned about their privacy disclosed less information (_r_ `r parameterestimates(fit_baseline, standardized = TRUE) %>%  filter(op == "~~" & lhs == "pri_con" & rhs == "Words_log") %>% select(std.all) %>%  my_round("std_txt")`). 
Specific gratifications predicted communication better than general gratifications (_r_ `r parameterestimates(fit_baseline, standardized = TRUE) %>%  filter(op == "~~" & lhs == "grats_spec" & rhs == "Words_log") %>% select(std.all) %>%  my_round("std_txt")` vs. _r_ `r parameterestimates(fit_baseline, standardized = TRUE) %>%  filter(op == "~~" & lhs == "grats_gen" & rhs == "Words_log") %>% select(std.all) %>%  my_round("std_txt")`).
The mean of privacy deliberation was _m_ = `r des_pridel$m %>% round(2)`. 
Altogether, `r pridel_active %>% round(0)`% of participants reported having actively deliberated about their privacy.

>>>>

abstract          : "The privacy calculus model states that people weigh risks and benefits before communicating online. This paper analyzes three questions: How robust is the privacy calculus, can it be replicated with actual behavior in realistic online settings? Do we need to extend the model by integrating more socio-psychological measures such as trust, self-efficacy, and privacy deliberation? How strongly is the privacy calculus affected by affordances such as like and dislike buttons? In a preregistered one-week field experiment (N = 590), participants discussed a current political topic on an online website, which included either (a) like buttons, (b) both like and dislike buttons, or (c) no buttons. The privacy calculus model was confirmed: Benefits and concerns affected communication. Deliberating about privacy by comparing benefits and risks decreased communication, whereas experiencing self-efficacy and trust increased communication. Although like and dislike buttons did not affect concerns, experiences, trust, efficacy, and deliberation, they had a strong effect on behavior: They reduced communication by 45 percent."

@@

bibliography      :
  - [bibliography/r-references.bib]
  - [bibliography/references.bib]
floatsintext      : yes
figsintext        : yes
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : no
mask              : yes
wordcount         : "8,204"
draft             : no
output            : papaja::apa6_pdf
documentclass     : "apa7"
classoption       : "man"
header-includes:
  - |
    \makeatletter
    \renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
      {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
      {-1em}%
      {\normalfont\normalsize\bfseries\typesectitle}}

    \renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
      {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
      {-\z@\relax}%
      {\normalfont\normalsize\bfseries\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
    \makeatother
    \usepackage{endnotes}
    \setlength{\parskip}{0em}
    \raggedbottom
    \note{\clearpage}

csl               : "`r system.file('rmd', 'apa7.csl', package = 'papaja')`"
# output:
#   html_document:
#     toc: TRUE
#     toc_float:
#       collapsed: FALSE
#       toc_depth: 2
---

@@

invisible(
  lapply(
    pcks 
    # , install.packages
    , library
    , character.only = TRUE
    )
  )

@@

According to the privacy paradox, the way users share information online is erratic [@barnesPrivacyParadoxSocial2006]:
Although being strongly concerned about their privacy, people communicate much personal information online  [@taddickenUsesPrivacyOnline2011].
However, despite its popularity in both academic research [@masurUnderstandingEffectsConceptual2023] and the public press [@newyorkpublicradioPrivacyParadox2018], empirical support for the privacy paradox is tenuous [@baruhOnlinePrivacyConcerns2017]. 
Research is increasingly building on the privacy calculus model [@dienlinPrivacyCalculusTheories2023], which states that communication online can be explained by perceived risks and expected benefits [@bolUnderstandingEffectsPersonalization2018; @krasnovaOnlineSocialNetworks2010].

Although the privacy calculus has gained momentum in academic research, several important questions remain unanswered. 
First, the privacy calculus' empirical foundation needs to be improved [@knijnenburgDeathPrivacyCalculus2017]. 
To date, much research on the privacy calculus builds on (a) self-reports of behavior [@krasnovaOnlineSocialNetworks2010], (b) vignette approaches [@bolUnderstandingEffectsPersonalization2018], or (c) one-shot experiments in the lab [@treptePrivacyCalculusContextualized2020]. 
Self-reports are often unreliable measures of online behavior [@parrySystematicReviewMetaanalysis2021], vignette approaches have limited external validity, and one-shot experiments cannot analyze effects of sustained use.
The first aim is therefore to replicate the privacy calculus model in a more authentic and long-term setting using actual behavioral data.
Doing so will show its robustness, and it will help understand if behavior is better explained by the privacy paradox approach or by the privacy calculus model.

Second, the privacy calculus model is criticized for neglecting the actual process of weighing pros and cons, and for over-emphasizing rationality  [@knijnenburgDeathPrivacyCalculus2017].
According to critics, simply showing that both concerns and gratifications correlate with communication online does not prove that an explicit weighing process took place.
Instead, we would need to find out if people deliberately compare costs and benefits before communicating online. 
Therefore, I will also analyze the process of comparing pros and cons explicitly, by elaborating on what I will call the the privacy deliberation process [@omarzuDisclosureDecisionModel2000]. 
Next, focusing exclusively on costs and benefits might not fully capture how and why users engage online.
I, hence, extend the privacy calculus by analyzing how online communication is affected by trust and self-efficacy, two variables less focused on rationality and more oriented toward socio-psychological aspects critical in online contexts [@metzgerPrivacyTrustDisclosure2004].

Online behavior is determined by the person itself and their external circumstances [@bazarovaIntegrationIndividualisticNetworked2020; @treptePrivacyCalculusContextualized2020; @spottswoodShouldShareThat2017].
The privacy calculus upholds that much of behavior is rational and self-determined, but how strongly does this process depend on online affordances?
It is well known that the affordances of many online services are optimized to elicit as much interaction as possible [@ellisonSocialNetworkSite2015; @masurBehavioralContagionSocial2021], for example via low threshold communication features such as likes, shares, replies, or reactions  [@carrPredictingThresholdPerceived2018]. 
Implicit and explicit cues on how a website is used can increase communication [@spottswoodShouldShareThat2017; @treptePrivacyCalculusContextualized2020].
How large are the effects of external popularity cues such as like and dislike buttons compared to the internal weighing of pros and cons?
And how easily can the mechanisms of the privacy calculus be affected by means of popularity cues?

In conclusion, in this paper we contribute to communication theory by means of (a) replication, (b) falsification, (c) extending the range of existing theory, (d) elucidating mechanisms, and (e) theory comparison [@deandreaIncreasingClarityWhere2017].

@@

The privacy calculus analyzes why people communicate online.
When are we willing to engage in a conversation?
It builds on the calculus of behavior [@lauferPrivacyConceptSocial1977] and states that people are weighing risks and benefits before actively communicating.
Communication is closely related to privacy, which is defined as a "voluntary and temporary withdrawal of a person from the general society" [@westinPrivacyFreedom1967, p. 7].
People regulate their privacy by deciding what and what not to communicate [@dienlinPrivacyProcessModel2014].
Communication is therefore also closely related to self-disclosure.
Just as it is impossible to not communicate, communication inherently promotes self-disclosure. 
In a recent study, communication quantity and the frequency of expressing one's political opinion was almost indistinguishable (_r_ = .91; AUTHORS).
Next to breadth and depth, communication quantity is hence often considered a central dimension of self-disclosure [@omarzuDisclosureDecisionModel2000].

Sharing information carries risks, as recipients may reuse it in different contexts, potentially harming the original sender [@masurBehavioralContagionSocial2021].
As a result, people who aim to avoid risks should be inclined to share less information, especially in online contexts where audiences are much larger [@vitakImpactContextCollapse2012].
Indeed, empirical research confirms that people with higher privacy needs and privacy concerns are less likely to communicate online [@krasnovaOnlineSocialNetworks2010; @masurTransformativeNotHow2021; @masurUnderstandingEffectsConceptual2023].
This finding is now confirmed by several meta-analyses and reviews [@baruhOnlinePrivacyConcerns2017; @dienlinDoesPrivacyParadox2021; @gerberExplainingPrivacyParadox2018].
Similarly, people who are more concerned about their privacy also engage in more privacy protection behaviors [@baruhOnlinePrivacyConcerns2017; @stubenvollLivingEasyEyes2022].

Even outweighing concerns, the most relevant drivers of online communication are expected gratifications [@dienlinExtendedPrivacyCalculus2016; @bolUnderstandingEffectsPersonalization2018; @kezerGettingPrivacyCalculus2022].
On average, people are happy to trade in parts of their privacy to receive something more valuable in return [@lauferPrivacyConceptSocial1977].
In online communication, the most important benefits include social support, social capital, entertainment, information-seeking, and self-presentation [@dhirUnderstandingRelationshipIntensity2017; @ellisonBenefitsFacebookFriends2007; @krasnovaOnlineSocialNetworks2010; @whitingWhyPeopleUse2013].

@@

Although privacy calculus implies that people explicitly compare benefits and risks before communicating online, prior research has neglected this aspect [@knijnenburgDeathPrivacyCalculus2017].
Only observing that privacy concerns or expected gratifications and communication online are related does not prove that an explicit weighing process took place [@knijnenburgDeathPrivacyCalculus2017]. 
Instead, we need to analyze if, and if so by how much, people actively deliberate about their privacy by comparing benefits and risks, and whether doing so influences their willingness to communicate.
Self-disclosure theory [@omarzuDisclosureDecisionModel2000;  @altmanPrivacyConceptualAnalysis1976] suggests that if the benefits of communication are attractive, deciding whether or not to communicate is a "conscious and deliberative process"  [@omarzuDisclosureDecisionModel2000, p. 183].
I hence introduce and investigate a novel concept termed _privacy deliberation_.
Privacy deliberation captures the extent to which individuals explicitly compare risks and benefits before communicating with others.

@@

<!-- The aforementioned weighing process is not flawless or perfect. -->
It is useful to understand the privacy calculus from the perspective of _bounded rationality_ [@simonBoundedRationality1990].
Bounded rationality states that "(1) humans are cognitively constrained; (2) these constraints impact decision making; and (3) difficult problems reveal the constraints and highlight their significance." [@bendorBoundedRationality2015, p. 1303].
It is important to emphasize that although human behavior is considered _partly_ irrational, bounded rationality does not state that it is _completely_ irrational [@gigerenzerBoundedRationalityAdaptive2002].
Instead, rationality needs to be understood as a continuum. 
And in the context of online privacy, rationality is impeded by information asymmetries, presence bias, intangibility, illusory control, or herding [@acquistiSecretsLikesDrive2020].
It follows that to provide a more complete picture, additional factors less focused on rationality but more on socio-psychological aspects should also explain communication.

Two central factors that help us understand online communication are self-efficacy and trust [@hossainICTEnabledVirtual2004; @metzgerPrivacyTrustDisclosure2004].
Privacy violations create psychological distress [@ledbetterParentchildPrivacyBoundary2019].
Experiencing online contexts as a safe space that users can sufficiently control is important for engaging in online communication. 
If users are more familiar, experienced, and knowledgeable in a given online context, they are more likely to navigate that online contexts successfully and to communicate actively [@baruhOnlinePrivacyConcerns2017; @kramerMasteringChallengeBalancing2020; @parkDigitalLiteracyPrivacy2013a].
People with more privacy self-efficacy engage more successfully in self-withdrawal [@dienlinExtendedPrivacyCalculus2016] and protective online behavior [@boermanExploringMotivationsOnline2021a; @vanooijenPrivacyCynicismIts2022].
Hence, if users possess more self-efficacy to participate, they should also communicate more.

@@

In all situations where people lack experience, control, or competence, a central variable to understand behavior is trust [@gefenTrustTAMOnline2003].
Trust plays a key role especially in online contexts [@metzgerPrivacyTrustDisclosure2004; @metzgerEffectsSiteVendor2006a].
Users often cannot control the environment or the way their information is handled [@acquistiSecretsLikesDrive2020; @braunlichLinkingLooseEnds2020].
Trust either captures "_specific_ beliefs dealing primarily with the integrity, benevolence, and ability of another party" [@gefenTrustTAMOnline2003, p. 55, emphasis added] or a "_general_ belief that another party can be trusted" [@gefenTrustTAMOnline2003, p.55, emphasis added]. 
In online contexts, there are different targets of trust, including (a) the information system, (b) the provider, (c) the Internet, and (d) the community of other users [@sollnerWhyDifferentTrust2016]. 
People who put more trust in the providers of networks, for example, disclose more personal information [@liEmpiricalStudiesOnline2011]. 
To comprehensively capture and understand online communication, trust should hence be included.

@@

How are the privacy calculus and communication affected by the context, the digital infrastructure? 
How easily can the calculus be manipulated externally?
One of the central tools to afford and govern online behavior are popularity cues such as like and dislike buttons [@stsiampkouskayaNotExperimentalStudy2023].
Popularity cues have been shown to affect behavior [@kramerMasteringChallengeBalancing2020; @masurBehavioralContagionSocial2021; @treptePrivacyCalculusContextualized2020].
For example, online comments that already have several dislikes are more likely to receive further dislikes [@muchnikSocialInfluenceBias2013]. 
When users disagree with a post, they are more likely to click on a button labeled _respect_ compared to a button labeled _like_ [@stroudRecommendRespectAltering2017].

How and why might popularity cues affect the privacy calculus?
In analyzing this question, it makes sense to analyze the cues' underlying affordances [@ellisonSocialNetworkSite2015; @foxDistinguishingTechnologiesSocial2017].
Affordances are mental representations of how objects might be used.
They emphasize that it is not the _objective features_ that determine behavior, but rather our _subjective perceptions_ [@gibsonEcologicalApproachVisual2015].
Popularity cues such as like and dislike buttons, which are of primary interest in this study, are understood as "paralinguistic digital affordances" [@carrPredictingThresholdPerceived2018, p. 142], lowering thresholds to partake in online communication.

Popularity cues likely impact the privacy calculus via two underlying theoretical mechanisms [@kramerMasteringChallengeBalancing2020]:
First, the _mere presence_ of popularity cues might affect whether people are willing to disclose. 
Being able to attract likes might afford active communication. 
The mere option to receive dislikes, conversely, might inhibit communication.
Second, _actually receiving_ likes or dislikes might affect behavior, by means of positive reinforcement (likes) or punishment (dislikes) [@skinnerScienceHumanBehavior2014].
To illustrate, likes are affirmative and embody the positivity bias of social media [@schreursExposurePositivityBias2022]. 
Receiving a like online is similar to receiving a compliment offline [@carrPredictingThresholdPerceived2018; @sumnerFunctionalApproachFacebook2017].
Like buttons afford and emphasize a _gain frame_ [@rosoffHeuristicsBiasesCyber2013].
These gains can be garnered only through active participation. 
In situations where people can gain immediate positive outcomes, concerns and risks that are more vague and in the future often become less relevant [presence bias, @ainslieHyperbolicDiscounting1992].
Because like buttons emphasize positive outcomes, it is likely that concerns decrease. 
In situations where there is more to win, people might also more actively deliberate about whether or not to communicate.

Dislikes, instead, represent a punishment, introducing a _loss frame_. 
Websites featuring both like _and_ dislike buttons should therefore be more ambivalent compared to websites without popularity cues, fostering privacy deliberation.
Privacy concerns should not be reduced anymore:
People who need more privacy are also more shy and risk averse [@dienlinWhoNeedsPrivacy2024].
Implementing the dislike button might therefore increase privacy concerns, canceling out the positive effects of the like button.
At the same time, communication and benefits might still be increased compared to a website without like and dislikes buttons, as online benefits are often considered to outweigh risks [positivity bias, @schreursExposurePositivityBias2022].

@@

In conclusion, this leads to an updated and extended model of the privacy calculus, in which the affordances of online contexts affect communication mediated via experienced privacy concerns, expected gratifications, privacy deliberation, self-efficacy, and trust. For an overview of the theoretical model, see Figure \@ref(fig:model).

```{r model, fig.cap="Overview of the extended privacy calculus model.", fig.align='center', out.width=".8\\textwidth"}
# knitr::include_graphics("figures/design/model_simplified.pdf")
knitr::include_graphics("figures/design/model_simplified.png")
```

@@

The data were collected in Germany.
Participants were recruited using the professional panel agency Norstat.
As incentive, participants were awarded digital points, to receive special offers from online retailers. 
Participants had to be over 18 years old and reside in Germany.
In a first step, the company sent its panel members an invitation to participate in the study. 
In this invitation, panel members were asked to participate in a study analyzing the current threat posed by terrorist attacks.
Members who agreed to participate were sent the first questionnaire (_T1_).
The questionnaire asked participants about their sociodemographic background, (b) provided more details about the study, and (c) included a registration link for the website, which was introduced as "participation platform".
Afterward, participants were randomly assigned to one of the three websites. 
After registration was completed, participants were invited (but not obliged) to visit the website and to discuss the topic of the terrorism threat in Germany. 
Participants could use the website and write comments over the course of one week.
Subsequently, participants received a follow-up questionnaire in which the self-reported measures reported below were collected (_T2_).

@@

```{r CFA, results = "asis"}
apa_table(
  facval_tab %>% 
    select(-alpha) %>% 
    filter(rownames(.) != "Gratifications specific") %>% 
    rename(`p-value` = pvalue) %>% 
    mutate(df = as.integer(.$df)), 
  font_size = "footnotesize",
  caption = "Psychometric Properties, Factorial Validity, and Reliability of Measures",
  note = "omega = Raykov's composite reliability coefficient omega; avevar = average variance extracted."
  , align = c("l", rep("r", 11))
  )
```

@@

_General gratifications_ were measured with five items based on @sunLocationInformationDisclosure2015.
One example item was "Using the participation platform has paid off for me". 
The mean was _m_ = `r facval_tab["Gratifications", "m"]`, which was above the scale's midpoint.
This shows that on average people considered the website to be beneficial.
For exploratory analyses, we also collected a scale measuring specific gratifications, not included here.

@@

I first plotted the bivariate relations of all variables (see Figure \@ref(fig:corrplot)). 
All variables referring to the privacy calculus demonstrated the expected bivariate relationships with communication. 
For example, people who were more concerned about their privacy disclosed less information (_r_ `r parameterestimates(fit_baseline, standardized = TRUE) %>%  filter(op == "~~" & lhs == "pri_con" & rhs == "Words_log") %>% select(std.all) %>%  my_round("std_txt")`). 
The mean of privacy deliberation was _m_ = `r des_pridel$m %>% round(2)`. 
Altogether, `r pridel_active %>% round(0)`% of participants reported having actively deliberated about their privacy.

The bivariate results showed two large correlations: 
specific trust and gratifications (_r_ = `r d_fs %>% select("Trust\nspecific", "Gratifications") %>% cor %>% .[2, 1] %>% my_round("std")`), and
privacy concerns and privacy deliberation (_r_ = `r d_fs %>% select("Privacy\nconcerns", "Privacy\ndeliberation") %>% cor %>% .[2, 1] %>% my_round("std")`). 
As all six variables were later analyzed within a single multiple regression, problems of multicollinearity might occur.
